---
title: "Raster Time-Series"
output:
  html_notebook: default
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---


This tutorial covers how to work with and plot raster time series data. The code and example datasets in this notebook were adapted from [NEON](http://www.neonscience.org/raster-time-series) data tutorials. It also covers practical assessment of data quality in remote sensing derived imagery.

##Downloading Landsat Imagery
For this tutorial we'll use a new GEE based web app for aquiring Landsat composit imagery (i.e., LS 5-8).
The web app is found [here](https://ndvi.ntsg.umt.edu/).
See the associated paper for an in depth description of processing method: [Robinson et al. 2017](http://www.mdpi.com/2072-4292/9/8/863)

##R Code
Install packages if not already installed
```{r}
# 
required.packages <- c("raster", "rgdal", "rasterVis", "ggplot2", "MODIS", "zoo", "imputeTS", "signal")
new.packages <- required.packages[!(required.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(required.packages, require, character.only=T)
rm(required.packages, new.packages)
```

Set working directory to ensure R can find the file we wish to import
```{r}
setwd("D:/EPA_Data/Master_File/Work_Files/Teaching/DataClass_TimeSeries")
```

##Working with NDVI Timeseries from southern NM

Create list of NDVI file paths
```{r}
IBPE_NDVI_path <- paste0(getwd(), "/IBPE") 
all_NDVI_IBPE <- list.files(IBPE_NDVI_path,
                            full.names = TRUE,
                            pattern = ".tif$")
# view list - note the full path, relative to our working directory, is included
print(all_NDVI_IBPE)
```
use gdalinfo to examine one of the geotiffs
```{r}
GDALinfo(all_NDVI_IBPE[1])
```

Now we have a list of all GeoTIFF files in the NDVI directory for IBPE site. Next, we will create a Rasterbrick from this list using the brick() function.

```{r}
# Create a raster brick of the NDVI time series
NDVI_IBPE_brick <- brick(stack(all_NDVI_IBPE, bands=1))
NDVI_QA_IBPE_brick <-  brick(stack(all_NDVI_IBPE, bands=2))
```

We can explore the GeoTIFF tags (the embedded metadata) in a brick using the same syntax that we used on single-band raster objects in R including: crs() (coordinate reference system), extent() and res() (resolution; specifically yres() and xres()).
```{r}
## ----explore-Rasterbrick-tags--------------------------------------------
# view crs of rasters
crs(NDVI_IBPE_brick)

# view extent of rasters in brick
extent(NDVI_IBPE_brick)

# view the resolution of our rasters
res(NDVI_IBPE_brick)
```

Data are scaled by 100. 
```{r}
NDVI_IBPE_brick <- NDVI_IBPE_brick/100
plot(NDVI_IBPE_brick[[1:12]], zlim=c(.06, .19), nc=4)
```


##LS QA
The quality band has the following values:
10: Pixel composite excludes clouds, cloud shadow, water, and snow
11: Pixel composite excludes clouds, cloud shadow, water, and snow; and is smoothed
20: Pixel composite excludes clouds and cloud shadow, but includes water and snow
21: Pixel composite excludes clouds and cloud shadow, but includes water and snow; and is smoothed
30: Pixel is filled with the user specified climatology
31: Pixel is filled with the user specified climatology; and is smoothed
255: No data

```{r}
plot(NDVI_QA_IBPE_brick[[1:12]])
```



```{r}
# using projectExtent (no values are transferred)
NDVI_IBPE_newproj <- projectExtent(NDVI_IBPE_brick,  "+proj=utm +zone=13 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
# Adjust the cell size 
res(NDVI_IBPE_newproj) <- 30
# now project
NDVI_IBPE_brick_proj <- projectRaster(NDVI_IBPE_brick, NDVI_IBPE_newproj)
NDVI_QA_IBPE_brick_proj <- projectRaster(NDVI_QA_IBPE_brick, NDVI_IBPE_newproj)
```


```{r}
## ----explore-Rasterbrick-tags--------------------------------------------
# view crs of rasters
crs(NDVI_IBPE_brick_proj)

# view extent of rasters in brick
extent(NDVI_IBPE_brick_proj)

# view the resolution of our rasters
res(NDVI_IBPE_brick_proj)
```

In the above exercise, we viewed plots of our NDVI time series and noticed a few images seem to be unusually light. However this was only a visual representation of potential issues in our data. Next we will use histograms to explore the distribution of NDVI values stored in each raster. 

```{r}
# create histograms of each raster
hist(NDVI_IBPE_brick_proj[[1:12]], 
     xlim = c(0, .4))
hist(NDVI_IBPE_brick_proj[[13:24]], 
     xlim = c(0, .4))
hist(NDVI_IBPE_brick_proj[[25:36]], 
     xlim = c(0, .4))
hist(NDVI_IBPE_brick_proj[[37:46]], 
     xlim = c(0, .4))
```



##Extracting date information

Print the basename of the first file to examine naming structure
```{r}
print(basename(all_NDVI_IBPE[1]))
```


The 'orgTime' function from the MODIS package lets you define the period to be filtered, the output temporal resolution, and select the required data from your input 'files'.
```{r}
date.org <- orgTime(all_NDVI_IBPE,nDays=16, begin="2009001", pillow=0, pos1=19,pos2=25)
print(date.org)
```

Select correct dates and assign date to each layer in the raster brick
```{r}
#Extract the formated input dates
ibpe.ls.dates <- date.org$inputLayerDates
#add timestamp to raster brick
NDVI_IBPE_brick_proj <- setZ(x = NDVI_IBPE_brick_proj, z = ibpe.ls.dates, 'date')
NDVI_QA_IBPE_brick_proj <- setZ(x = NDVI_QA_IBPE_brick_proj, z = ibpe.ls.dates, 'date')
```


Here we will us both the zoo and raster packages to process our time-series raster stack

Create quartly NDVI averages
```{r}
NDVI_IBPE_yearqrt <- zApply(NDVI_IBPE_brick_proj, by=as.yearqtr, fun=mean, name='quarters')
#plot avg quarterly values
plot(NDVI_IBPE_yearqrt)
```

Create monthly NDVI averages
```{r}
NDVI_IBPE_yearmon <- zApply(NDVI_IBPE_brick_proj, by=as.yearmon, fun=mean, name='months')
#plot avg monthly values in 2009
plot(NDVI_IBPE_yearmon[[1:12]])
```

We can use the rasterVis package for plotting
```{r}
cols <- colorRampPalette(brewer.pal(9,"YlGn"))
levelplot(NDVI_IBPE_yearmon,
          layout=c(4, 3), # create a 4x4 layout for the data
          col.regions=cols, # add a color ramp
          main="Landsat NDVI Monthly Average")
```

Now we will load in some points to extract timeseries from the raster brick
```{r}
points <- readOGR("D:/EPA_Data/Master_File/Work_Files/Teaching/DataClass_TimeSeries", "IBPE_Points")
ts.points <- extract(NDVI_IBPE_brick_proj, points)
plot(NDVI_IBPE_brick_proj[[1]])
points(points, pch=16)
```

#Based on areas on highest temporal variability we will select two points to evaluate
```{r}
plot(NDVI_IBPE_brick_proj[[15]])
points(points[1,], pch=16, col="red")
points(points[5,], pch=16, col="blue")
```

Here we plot and compare the points
```{r}
p1 <- zoo(ts.points[1,],ibpe.ls.dates)
autoplot.zoo(p1)
p5 <- zoo(ts.points[5,],ibpe.ls.dates)
autoplot.zoo(p5)
```

Lets dig further into the QA data
##LS QA
The quality band has the following values:
10: Pixel composite excludes clouds, cloud shadow, water, and snow
11: Pixel composite excludes clouds, cloud shadow, water, and snow; and is smoothed
20: Pixel composite excludes clouds and cloud shadow, but includes water and snow
21: Pixel composite excludes clouds and cloud shadow, but includes water and snow; and is smoothed
30: Pixel is filled with the user specified climatology
31: Pixel is filled with the user specified climatology; and is smoothed
255: No data

```{r}
ts.qa.points <- extract(NDVI_QA_IBPE_brick_proj, points)
p1.qa <- zoo(ts.qa.points[1,],ibpe.ls.dates)
autoplot.zoo(p1.qa)
p5.qa <- zoo(ts.qa.points[5,],ibpe.ls.dates)
autoplot.zoo(p5.qa)
```

```{r}
p1_int <- p1
p5_int <- p5
```

```{r}
p1[p1.qa==30] <- NA
#Note: for some reason p5.qa is not returning values == 30. p1 and p5 have the same QA so I used p1 to select infilled values for p5.
p5[p1.qa==30] <- NA
autoplot.zoo(p1)
autoplot.zoo(p5)
```

The zoo and imputeTS packages has several functions for interpolating missing values in a timeseries
```{r}
#Zoo package
p1.napprox <- na.approx(p1)
p1.naspline <- na.spline(p1)

#imputeTS package
p1.linear <-zoo(na.interpolation(as.numeric(p1), option = "linear"),ibpe.ls.dates)
p1.spline <-zoo(na.interpolation(as.numeric(p1), option = "spline"),ibpe.ls.dates)
p1.stine <-zoo(na.interpolation(as.numeric(p1), option = "stine"),ibpe.ls.dates)

autoplot.zoo(p1)
autoplot.zoo(p1.napprox)
autoplot.zoo(p1.naspline)
autoplot.zoo(p1.linear)
autoplot.zoo(p1.spline)
autoplot.zoo(p1.stine)
autoplot.zoo(p1_int)
```

```{r}
save.image(paste0(getwd(), "/DataClass_TimeSeries.RData"))
```

