---
title: "Iteration and Looping"
output: html_notebook
---

Iteration and looping in R is useful for when you need to do the same operation over a set of objects, datasets, or observations, or when you want to run some simulations. You can easily set up iteration in R over multiple variables (as long as you can keep your head around that they're doing!). The basic iterator in R is the __for()__ function

Check out this example of the basic for loop syntax.
```{r}
library(dplyr)

for (i in 1:10) {  # Basic structure 
  print(i)        # i is the "iterator". It's a counter in this case.
}
```
 You can also iterate over lists of objects
 
Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.
```{r}
for (month in c("Jan","Feb","Mar","Apr","May")) {
  print(month)
}
```

Use different R functions to create sets of objects to iterate over
```{r}
for (rand.num in rnorm(12)) {
  print(paste("The Random number is: ",rand.num))
}
```

Grab a list of files from a directory and iterate over those
```{r}
data.path <- "C:\\Users\\Jason Karl\\Documents\\GitHub\\REM504-DataScience\\Practice_Datasets"
files <- list.files(data.path,pattern="*.csv")
i <- 1 # set up a counter here so we can load the datasets iteratively
out.data <- c()
for (file in files) {
  print(file)
  out.data[[i]]<-read.csv(paste(data.path,file,sep="\\"))
  i<-i+1
}
```

For simulations, you can use loops to build your results
```{r}
out.data <- data.frame() # set up an empty data frame
for (i in 1:10) {
  var.rand <- rnorm(1)
  var.now <- date()
  var.other <- "Jason"
  out.row <- data.frame(i=i, rand=var.rand, now=var.now, me=var.other)
  out.data <- rbind(out.data,out.row)
}
glimpse(out.data)
```


You can also nest for loops
```{r}
out.data <- data.frame() # set up an empty data frame
for (i in 1:10) {
  for (j in 1:15) {
    var.rand <- rnorm(1)
    var.now <- date()
    var.other <- "Jason"
    out.row <- data.frame(i=i, j=j, rand=var.rand, now=var.now, me=var.other)
    out.data <- rbind(out.data,out.row)
  }
}
glimpse(out.data)
```

